Galaxy Imaging with Generative Models: Insights from a Two-Models Framework

### 2025-03-13 MNRAS/galaxy-generative-models-overleaf-submit-2nd/article_MNRAS.tex
dataset section

% Abstract of the paper
\begin{abstract}
Generative models have recently revolutionized image generation tasks across diverse domains, including galaxy image synthesis. This study investigates the statistical learning and consistency of three generative models: \texttt{light-weight-gan} (a GAN-based model), \texttt{Glow} (a Normalizing Flow-based model), and a diffusion model based on a \texttt{U-Net} denoiser, all trained on non-overlapping subsets of the SDSS DR7 dataset of $64 \times 64$ grayscale images. While all models produce visually realistic images with well-preserved morphological variable distributions, we focus on their ability to learn and generalize the underlying data distribution.

@@

To achieve this, \cite{kadkhodaie2024generalization} employed \textit{denoiser} architectures, such as \texttt{U-Net} \citep{ronneberger2015u} and \texttt{BF-CNN} networks \citep{Mohan2020Robust}, trained on reduced datasets like \texttt{CelebA} \citep{Liu2015} and \texttt{LSUN} Bedroom \citep{Yu2015}, comprising $O(10^5)$ grayscale images resized to $80 \times 80$ pixels.  This dataset scale enables a systematic investigation of how various generative models, including normalizing flows, GANs, and denoiser-based diffusion models, perform in galaxy image generation when trained on the same dataset derived from the SDSS DR7 survey.

@@

{\color{red}
\subsection{Some comments}
%

@@

{\color{red}
In the present study, we focus on galaxy image generation, having as underlying guide the experiment performed by \cite{kadkhodaie2024generalization} who  employed widely used machine learning datasets. All images were grayscale (single channel) with intensity values in the range $[0,1]$ and resolution of $80 \times 80$ pixels. The transition from memorization to generalization was observed with a training dataset consisting of $10^5$ images.

Given the requirement to train models with at least $10^5$ images, we used the Sloan Digital Sky Survey (SDSS) dataset described in \citep{smith2021}\footnote{Raw data available at \url{https://github.com/Smith42/astroddpm}.} with modifications to preprocessing. Specifically, all images were cropped around target galaxy coordinates to $64 \times 64$ pixels, and a single grayscale channel was created from the $(z, g, r)$ channels using the \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{The \texttt{make\_lupton\_rgb} function was applied with $Q=8$ and \texttt{stretch}=0.2 settings.}. 
}
%
\subsection{The models}
\label{sec-Exp-Models}
%
We selected three architectures, as described in Section~\ref{sec-generative-models}: a GAN, a flow-based model, and a diffusion-based model. We have used implementations of architectures written in \texttt{PyTorch} \citep{PyTorch2019}.

@@

For the diffusion-based model, we adopted a \texttt{U-Net} architecture \citep{ronneberger2015u} as the \textit{denoiser}, leveraging its multi-scale structure with $7.6$ million parameters\footnote{Source code: \url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}.}. Training followed the procedure in \cite{kadkhodaie2024generalization}, with $100$ epochs taking about $100$ hours.
%
\subsection{Results}
\label{sec-Results}
%
\subsubsection{Generated images}
\label{sec-Generated-Images}
%
Figure~\ref{fig-Original-Glow-UNet-Gan-samples} shows examples of galaxy images (\textit{validation}) from the original dataset alongside samples generated by the three different models trained with $N = 10^5$ images.
%
%
At first glance, the generated images appear indistinguishable from the original images. To provide a more quantitative evaluation, Figure~\ref{fig-morpho-coeff} presents various morphological coefficients computed using \texttt{Statmorph}\footnote{\url{https://github.com/vrodgom/statmorph}} \citep{2019MNRAS.483.4140R}, including Concentration, Asymmetry, and Smoothness \citep{2000AJ....119.2645B,2003ApJS..147....1C,2004AJ....128..163L}, as well as Gini and $M_{20}$ coefficients \citep{2004AJ....128..163L,10.1093/mnras/stv2078}.

>>>>

% Abstract of the paper
\begin{abstract}
Generative models have recently revolutionized image generation tasks across diverse domains, including galaxy image synthesis. This study investigates the statistical learning and consistency of three generative models: \texttt{light-weight-gan} (a GAN-based model), \texttt{Glow} (a Normalizing Flow-based model), and a diffusion model based on a \texttt{U-Net} denoiser, all trained on non-overlapping subsets of the SDSS {\color{red}DR9} dataset of $64 \times 64$ grayscale images. While all models produce visually realistic images with well-preserved morphological variable distributions, we focus on their ability to learn and generalize the underlying data distribution.

@@

To achieve this, \cite{kadkhodaie2024generalization} employed \textit{denoiser} architectures, such as \texttt{U-Net} \citep{ronneberger2015u} and \texttt{BF-CNN} networks \citep{Mohan2020Robust}, trained on reduced datasets like \texttt{CelebA} \citep{Liu2015} and \texttt{LSUN} Bedroom \citep{Yu2015}, comprising $O(10^5)$ grayscale images resized to $80 \times 80$ pixels.  This dataset scale enables a systematic investigation of how various generative models, including normalizing flows, GANs, and denoiser-based diffusion models, perform in galaxy image generation when trained on the same dataset derived from the SDSS {\color{red} DR9} survey.

@@

{\color{red}
\subsection{Some comments}
\label{sec-some-comments}
%

@@

%
\subsection{The models}
\label{sec-Exp-Models}
%
We selected three architectures, as described in Section~\ref{sec-generative-models}: a GAN, a flow-based model, and a diffusion-based model. We have used implementations of architectures written in \texttt{PyTorch} \citep{PyTorch2019}.

@@

For the diffusion-based model, we adopted a \texttt{U-Net} architecture \citep{ronneberger2015u} as the \textit{denoiser}, leveraging its multi-scale structure with $7.6$ million parameters\footnote{Source code: \url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}.}. Training followed the procedure in \cite{kadkhodaie2024generalization}, with $100$ epochs taking about $100$ hours.

{\color{red}
In the present study, we focus on galaxy image generation, taking as a reference the experiment conducted by \cite{kadkhodaie2024generalization}, who, however, used widely known machine learning datasets. All their images were grayscale (single channel), with intensity values in the range $[0,1]$ and a resolution mostly of $80 \times 80$ pixels.

We use the Sloan Digital Sky Survey (SDSS) Data Release 9 dataset, as described in \citep{smith2021}\footnote{The script to download the original selected dataset is available at \url{https://github.com/Smith42/astroddpm/data/sdss}.}. As the authors mentioned, the volume complete sample includes $(g,r,z)$ bands, with an $r$-band absolute magnitude limit of $M_r\leq 20$ and a redshift limit of $z\leq 0.08$. Each band image consists of $512\times 512$ pixels at a resolution of $0.262^{\prime\prime} \mathrm{pixel}^{-1}$ (aka \texttt{pixscale}), with an average seeing of approximately $1^{\prime\prime}$. Our preprocessing applied to the downloaded dataset, consisting of approximately $250,000$ images, includes cropping around the target galaxy coordinates to $64 \times 64$ pixels to ensure that each image predominantly contains a single object. Additionally, a single grayscale channel was created from the $(z, g, r)$ bands using the \texttt{scipy} \citep{2020SciPy-NMeth} library to balance the contributions of the different bands\footnote{The \texttt{make\_lupton\_rgb} function was applied with settings $Q=8$ and \texttt{stretch}=0.2.}. Of course, this choice of dataset is by no means restrictive, and researchers may use their own datasets. To facilitate reproducibility, we provide additional materials in the companion GitHub repository, enabling users to replicate the experiments.

Thus, we restrict ourselves to relatively modest image sizes compared to those produced by models such as \texttt{light-weight-gan} and \texttt{Glow}. Our primary objective is to investigate whether the transition from \textit{memorization} to \textit{generalization} in diffusion models occurs around $10^5$ training images, as mentioned earlier, and whether this transition aligns with the number of model parameters (see Section~\ref{sec-some-comments}). Furthermore, we aim to examine how GAN and flow-based models behave as a function of dataset size. However, even when using similar image sizes, it remains an open question whether the same number of training images is required to achieve the generalization transition when working with face or bedroom images compared to galaxy images. The complexity of such images can be estimated based on the compression ratio achieved using the same algorithm\footnote{For example, \texttt{JPEG} compression with default parameters in the \texttt{PIL} library v11.1.0 (\url{https://pillow.readthedocs.io/}).}. We find that the galaxy images used in this study exhibit a complexity reduction of approximately $50\%$. With less complex images, one might expect an easier task for generative models. At the very least, this question can be explored within the framework described here.
}

%
\subsection{Results}
\label{sec-Results}
%
\subsubsection{Generated images}
\label{sec-Generated-Images}
%
Figure~\ref{fig-Original-Glow-UNet-Gan-samples} shows examples of galaxy images (\textit{validation}) from the original dataset alongside samples generated by the three different models trained with $N = 10^5$ images.
%
%
At first glance, the generated images appear indistinguishable from the original images. To provide a more quantitative evaluation, Figure~\ref{fig-morpho-coeff} presents various morphological coefficients computed using \texttt{Statmorph}\footnote{\url{https://github.com/vrodgom/statmorph}} \citep{2019MNRAS.483.4140R}, including Concentration, Asymmetry, and Smoothness \citep{2000AJ....119.2645B,2003ApJS..147....1C,2004AJ....128..163L}, as well as Gini and $M_{20}$ coefficients \citep{2004AJ....128..163L,10.1093/mnras/stv2078}.
